\bf Estimating the number of clusters using cross-validation

### 2018-02-27 cvclust.tex
fix Ref 1 comments

A clustering procedure segments a collection of items into smaller groups,
with the property that items in the same group are more similar to each other
than items in different groups \citep{hartigan1975clustering}.  Such
procedures are used in two main applicaitons: (a) exploratory analysis, where clusters
reveal homogeneous sub-groups within a large sample; (b) data
reduction, where high-dimensional item attribute vectors get reduced to discrete
cluster labels \citep{jain1999data}.

@@

The problem of choosing $k$ has been well-studied, and dozens of methods have
been proposed~\citep{chiang2010intelligent,fujita2014non}. The main difficulty
in choosing $k$ is that clustering is fundamentally an ``unsupervised''
learning problem, meaning that there is no obvious way to use ``prediction
ability'' to drive the model selection \citep{hastie2009elements}.  Most
existing methods for choosing $k$ instead rely on explicit or implicit
assumptions about the data distribution, including it shape, scale, and
correlation structure.

@@

The procedure we propose in this report is based on a form of
cross-validation, and it is adaptive to the characteristics of the data
distribution. The essential idea is to devise a way to measure a form of
internal prediction error associated each choose of $k$, and then choose the
$k$ with the smallest associated error. We describe this method in detail in
Section~\ref{sec:meth}.  In Section~\ref{sec:self-consistent}, we prove that
our method is self-consistent. Then, in Section~\ref{sec:theory-gaussian}, we
analyze the performance of our method in the presence of Gaussian noise. The
theoretical analysis shows that the performance of our method degrades in the
presence of correlated noise; to fix this, we propose a correction for
correlated noise in Section~\ref{sec:corr-correct}. In
Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}, we
demonstrate that our method is competitive with other state-of-the-art
procedures in both simulated and real data sets. Then, in
Section~\ref{sec:application-yeast}, we apply our method to a Yeast cell cycle
dataset. We conclude with a short discussion in Section~\ref{sec:discussion}.

@@

\subsection{Problem statement}

Suppose that we are given a data matrix with $N$ rows and $P$ columns, and we
are tasked with choosing an appropriate number $k$ of clusters to use for
performing $k$-means clustering on the rows of the data matrix. Recall that
the $k$-means procedure takes a set of observations $\{ x_1, \dotsc ,x_n \}$
and finds a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
minimizing the within cluster dispersion
\[
  W(A) = \sum_{i=1}^{n} \min_{a \in A} \|x_i - a\|^2.
\]
This implicitly defines a cluster assignment rule
\[
  g(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
\]
with ties broken arbitrarily.
%We will assume that the $k$-means procedure
%finds an optimal choice of , $A$, but we will not assume that this solution is
%unique.

We can consider the problem of choosing $k$, the number of clusters, to be a
model selection problem. In other domains, especially supervised learning
problems like regression and classification, cross-validation is popular for
performing model selection.
In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  For each
``fold'' $r$ in the range $1, \dotsc, K$, we permute the rows of the data
matrix to get $\dataX$, a matrix with the $r$th test subset as its trailing
rows.  We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$ to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

Given such a partition of $\dataX$, we perform four steps for each value of
$k$, the number of clusters:
\begin{enumerate}
  \item \label{step:gabriel-cluster}
    \textbf{Cluster:}
    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding the
    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ and the
    cluster means $\bmuY_1, \dotsc, \bmuY_k$.  Set $\hGY_i = \hGY(Y_i)$ to
    be the assigned cluster for row $i$.
  \item \label{step:gabriel-classify}
    \textbf{Classify:}
    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to train a classifier
    $\hGX : \R^p \to \{ 1, \dotsc, k \}$.
  \item \label{step:gabriel-predict}
    \textbf{Predict:}
    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
    $\bmuY(g) = \bmuY_g$.
  \item \label{step:gabriel-evaluate}
    \textbf{Evaluate:}
    Compute the cross-validation error
    \[
      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
    \]
    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
\end{enumerate}
\noindent
In principle, we could use any clustering and classification methods in
steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
report, we use $k$-means \citep{hartigan1979algorithm} as the clustering algorithm 
and develop the theoretical properties of the proposed method based on $k$-means. 
For the classification step, we compute the mean value of $X$ for each class; we assign an
observation to class $g$ if that class has the closest mean (randomly breaking
ties between classes).  The classification step is equivalent to linear
discriminant analysis with equal class priors and identity noise covariance
matrix.

@@

In Section~\ref{sec:self-consistent}, we prove that this procedure is
self-consistent, in the sense that it recover the correct value of $k$ in the
absence of noise. Then, in Section~\ref{sec:theory-gaussian}, we analyze some
of the properties of Gabriel cross-validation in the presence of Gaussian
noise.

@@

An important property of any estimation procedure is that in the absence of of
noise, the procedure correctly estimates the truth. This property is called
``self-consistency'' \citep{tarpey96}. We will now show that Gabriel
cross-validation is self-consistent. That is, in the absence of noise, the
Gabriel cross-validation procedure finds the optimal number of clusters.

@@

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.
  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.
  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.
\end{enumerate}
Then $\CV(k) < \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
Gabriel cross-validation correctly chooses $k = K$.
\end{proposition}

>>>>

A clustering procedure segments a collection of items into smaller groups,
with the property that items in the same group are more similar to each other
than items in different groups \citep{hartigan1975clustering}.  Such
procedures are used in two main applications: (a) exploratory analysis, where clusters
reveal homogeneous sub-groups within a large sample; (b) data
reduction, where high-dimensional item attribute vectors get reduced to discrete
cluster labels \citep{jain1999data}.

@@

The problem of choosing $k$ has been well-studied, and dozens of methods have
been proposed~\citep{chiang2010intelligent,fujita2014non}. The main difficulty
in choosing $k$ is that clustering is fundamentally an ``unsupervised''
learning problem, meaning that there is no obvious way to use ``prediction
ability'' to drive the model selection \citep{hastie2009elements}.  Most
existing methods for choosing $k$ instead rely on explicit or implicit
assumptions about the data distribution, including its shape, scale, and
correlation structure.

@@

The procedure we propose in this report is based on a form of
cross-validation, and it is adaptive to the characteristics of the data
distribution. The essential idea is to devise a way to measure a form of
internal prediction error associated each choose of $k$, and then choose the
$k$ with the smallest associated error. We describe this method in detail in
Section~\ref{sec:meth}.  In Section~\ref{sec:self-consistent}, we prove that
our method is self-consistent in certain situations, meaning that in the
absence of noise, it recovers the correct $k$. Then, in
Section~\ref{sec:theory-gaussian}, we
analyze the performance of our method in the presence of Gaussian noise. The
theoretical analysis shows that the performance of our method degrades in the
presence of correlated noise; to fix this, we propose a correction for
correlated noise in Section~\ref{sec:corr-correct}. In
Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}, we
demonstrate that our method is competitive with other state-of-the-art
procedures in both simulated and real data sets. Then, in
Section~\ref{sec:application-yeast}, we apply our method to a Yeast cell cycle
dataset. We conclude with a short discussion in Section~\ref{sec:discussion}.

@@

\subsection{Problem statement}
\label{sec:problem-statement}

Suppose that we are given a data matrix with $N$ rows and $P$ columns, and we
are tasked with choosing an appropriate number $k$ of clusters to use for
performing $k$-means clustering on the rows of the data matrix. Recall that
the $k$-means procedure takes a set of observations $\{ x_1, \dotsc ,x_n \}$
and finds a set of $k$ or cluster centers $A = \{ a_1, \dotsc, a_k \} $
minimizing the within cluster dispersion
\[
    W(A) = \sum_{i=1}^{n} \min_{g \in \{ 1, \dotsc, k \} } \|x_i - a_g\|^2.
\]
This implicitly defines a cluster assignment rule
\[
  G(x) = \argmin_{g \in \{1, \dotsc, k\}} \|x - a_g\|^2,
\]
with ties broken arbitrarily.
%We will assume that the $k$-means procedure
%finds an optimal choice of , $A$, but we will not assume that this solution is
%unique.

We can consider the problem of choosing $k$, the number of clusters, to be a
model selection problem. In other domains, especially supervised learning
problems like regression and classification, cross-validation is popular for
performing model selection.
In these settings, the data comes in the form of $N$
predictor-response pairs, $(X_1, Y_1), \dotsc, (X_N, Y_N)$, with $X_i \in
\R^{p}$ and $Y_i \in \R^{q}$.  The data can be represented as a matrix with
$N$ rows and $p + q$ columns.  We partition the data into $K$ hold-out
``test'' subsets, with $K$ typically chosen to be $5$ or $10$.  We do this
by randomly assigning a fold number $r_i$ in the range $\{1, \dotsc, K\}$
for each row~$i$. For each ``fold'' $r$ in the range $1, \dotsc, K$, we
permute the rows of the data matrix to get $\dataX$, a matrix with the
$r$th test subset, the rows $i$ with $r_i = r$ as its trailing rows.  
We partition $\dataX$ as
\[
  \dataX =
  \begin{bmatrix}
    \Xtrain & \Ytrain \\
    \Xtest  & \Ytest
  \end{bmatrix}.
\]
We use the training rows $[ \Xtrain\ \Ytrain ]$, containing roughly $N
(K-1)/K$ of the observations to fit a regression model
$\hat Y = \hat Y(X)$, and then evaluate the performance of this model on the
test set, computing the cross-validation error $\|\Ytest - \hat Y(\Xtest)\|^2$
or some variant thereof.  We choose the model with the smallest
cross-validation error, averaged over all $K$ folds.

@@

Given such a partition of $\dataX$, we perform four steps for each value of
$k$, the number of clusters:
\begin{enumerate}
  \item \label{step:gabriel-cluster}
    \textbf{Cluster:}
    Cluster $Y_{1}, \dotsc, Y_n$, the rows of $\Ytrain$, yielding
    $k$ cluster centers, denoted $\bmuY_1, \dotsc, \bmuY_k$, identified
    with the labels $1, \dotsc, k$. This also yields the
    assignment rule $\hGY : \R^q \to \{ 1, \dotsc, k \}$ defined analogously
    to the assignment rule $G$ from Sec.~\ref{sec:problem-statement}, which
    for any input $Y$, yields the label of the closest cluster center.
    Set $\hGY_i = \hGY(Y_i)$ to be the assigned cluster label for row~$i$: the
    label of the closest cluster mean to $Y_i$.
  \item \label{step:gabriel-classify}
    \textbf{Classify:}
    Take $X_{1}, \dotsc, X_n$, the rows of $\Xtrain$ to be predictors,
    and take $\hGY_1, \dotsc, \hGY_n$ to be corresponding class labels.  Use
    the pairs $\{ (X_i, \hGY_i) \}_{i=1}^{n}$ to fit a model that can
    predicts a label from input vector $X$; denote the fitted
    prediction function by
    $\hGX : \R^p \to \{ 1, \dotsc, k \}$. In principle this prediction method
    can be arbitrary but in the sequel we will use linear discriminant
    analysis.
  \item \label{step:gabriel-predict}
    \textbf{Predict:}
    Apply the classifier to $X_{n+1}, \dotsc, X_{n+m}$, the rows of
    $\Xtest$, yielding predicted classes $\hGX_i = \hGX(X_i)$ for
    $i = n+1, \dotsc, n+m$.  For each value of $i$ in this range, compute
    predicted response $\hat Y_i = \bmuY(\hGX_i)$, where
    $\bmuY(g) = \bmuY_g$.
  \item \label{step:gabriel-evaluate}
    \textbf{Evaluate:}
    Compute the cross-validation error
    \[
      \CV(k) = \frac{1}{m} \sum_{i=n+1}^{n+m} \|Y_i - \hat Y_i\|^2,
    \]
    where $Y_{n+1}, \dotsc, Y_{n+m}$ are the rows of $\Ytest$.
\end{enumerate}
\noindent
In principle, we could use any clustering and classification methods in
steps~\ref{step:gabriel-cluster} and~\ref{step:gabriel-classify}.  In this
report, we use $k$-means \citep{hartigan1979algorithm} as the clustering algorithm 
and develop the theoretical properties of the proposed method based on $k$-means. 
For the classification step, we compute the mean value of $X$ for each class; we assign an
observation to class $g$ if that class has the closest mean (randomly breaking
ties between classes).  The classification step is equivalent to linear
discriminant analysis with equal class priors and identity noise covariance
matrix.

@@

In Section~\ref{sec:self-consistent}, we prove that this procedure is
self-consistent, in the sense that it recovers the correct value of $k$ in the
absence of noise. Then, in Section~\ref{sec:theory-gaussian}, we analyze some
of the properties of Gabriel cross-validation in the presence of Gaussian
noise.

@@

An important property of any estimation procedure is that in the absence of
noise, the procedure correctly estimates the truth. This property is called
``self-consistency'' \citep{tarpey96}. We will now show that Gabriel
cross-validation is self-consistent. That is, in the absence of noise, the
Gabriel cross-validation procedure finds the optimal number of clusters.

@@

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $\CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ ``true cluster centers''
$\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.
  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.
  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.
\end{enumerate}
Then $\CV(k) < \CV(K)$ for $k < K$, and $\CV(k) = \CV(K)$ for $k > K$, so that
Gabriel cross-validation correctly chooses $k = K$.
\end{proposition}
